Habitat edges affect tree diversity more than biomass regeneration in a reforested wet neotropical timber plantation

### 2022-11-09 article/merge.tex
bioRxiv,poster

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage[left]{lineno}
\linenumbers
\usepackage{setspace}\doublespace
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
\author{}
\date{\vspace{-2.5em}}

@@

*\textbf{Corresponding author}: Nicholas Medina, \href{mailto:nmedina@umich.edu}{\nolinkurl{nmedina@umich.edu}}

@@

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests.
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely.
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
This study addressed how wet tropical secondary forest regeneration was affected by distance to habitat edge, and hypothesized that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.

As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ca 10\% variation explained, while biomass and canopy light tended to stay the same.
Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably.
Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

@@

\hypertarget{abstracto}{%
\section*{Abstracto}\label{abstracto}}
\addcontentsline{toc}{section}{Abstracto}

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.

Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10\% variación explicado, mientras tanto biomasa y luz de dócel tendió a quedarse igual.
Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, pero biomasa por modo de disperso no tendió a cambiar.
Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.

@@

Forest and landscape restoration is a key international conservation and climate change adaptation strategy \emph{(\protect\hyperlink{ref-depinto20}{De Pinto et al., 2020})}, and while tropical forests specifically store most land biomass \emph{(\protect\hyperlink{ref-pan13}{Pan et al., 2013})}, most are now secondary \emph{(\protect\hyperlink{ref-fao20}{FAO, 2020})} and regenerating degraded functions \emph{(\protect\hyperlink{ref-hubau20}{Hubau et al., 2020})} amid increasing climate stressors \emph{(\protect\hyperlink{ref-anderegg22}{Anderegg et al., 2022})} yet insufficient policy support \emph{(\protect\hyperlink{ref-chazdon18}{Chazdon, 2018})} via skewed priorities \emph{(\protect\hyperlink{ref-pyron22}{Pyron and Mooers, 2022})}.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass \textgreater10x faster than old-growth forests \emph{(\protect\hyperlink{ref-finegan96}{Finegan, 1996}; \protect\hyperlink{ref-poorter16}{Poorter et al., 2016})}, and substantially offset carbon emissions, with estimates ranging from \textasciitilde10\% from Amazonian deforestation \emph{(\protect\hyperlink{ref-smith20}{Smith et al., 2020})} to a decade of fossil fuel emissions from all of Latin American and the Caribbean \emph{(\protect\hyperlink{ref-chazdon16}{Chazdon et al., 2016})}.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80\% of old-growth forest levels in just 20 years, especially in wetter regions \emph{(\protect\hyperlink{ref-rozendaal19}{Rozendaal et al., 2019})}.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services \emph{(\protect\hyperlink{ref-guariguata01}{Guariguata and Ostertag, 2001}; \protect\hyperlink{ref-stanturf14}{Stanturf et al., 2014})} like biomass and carbon storage \emph{(\protect\hyperlink{ref-liu18}{Liu et al., 2018})}.
However, secondary neotropical forests diverge widely in taxonomic composition \emph{(\protect\hyperlink{ref-norden15}{Norden et al., 2015})}, potentially taking over a century to recover \emph{(\protect\hyperlink{ref-poorter21}{Poorter et al., 2021})} with added variability \emph{(\protect\hyperlink{ref-atkinson22}{Atkinson et al., 2022})}.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa \emph{(\protect\hyperlink{ref-carlo16}{Carlo and Morales, 2016})}, which can be keystone to locally-adapted food webs, including birds \emph{(\protect\hyperlink{ref-maas16}{Maas et al., 2016})}.
Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, yet local management remains the strongest driver of tropical forest carbon stores \emph{(\protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}.

Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, as well as land-use history, especially in heterogeneously-managed landscapes \emph{(\protect\hyperlink{ref-melo13}{Melo et al., 2013}; \protect\hyperlink{ref-perfecto09}{Perfecto et al., 2009})}.
About 70\% of all forests have been estimated to be within a km of their edge \emph{(\protect\hyperlink{ref-haddad15}{Haddad et al., 2015})}, resulting in tropical forests having been estimated to lose \textasciitilde22\% aboveground carbon along heavily-managed edges \textgreater100m, in part due to shortening trees and lighter leaf mass \emph{(\protect\hyperlink{ref-ordway20}{Ordway and Asner, 2020})}.
Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa \emph{(\protect\hyperlink{ref-wandrag17}{Wandrag et al., 2017})}, which while complicated \emph{(\protect\hyperlink{ref-muller-landau09}{Muller-Landau and Hardesty, 2009})}, can also correlate with simpler plant traits like height and seed mass \emph{(\protect\hyperlink{ref-tamme14}{Tamme et al., 2014}; \protect\hyperlink{ref-thomson11}{Thomson et al., 2011})}.

Properties of the existing tree community also depend on land-use history and previous management \emph{(\protect\hyperlink{ref-omeja12}{Omeja et al., 2012})} including plantation use for timber production, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant \emph{(\protect\hyperlink{ref-tabarelli08}{Tabarelli et al., 2008})}.
Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages \emph{(\protect\hyperlink{ref-albrich20}{Albrich et al., 2020}; \protect\hyperlink{ref-gough21}{Gough et al., 2021})}, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry \emph{(\protect\hyperlink{ref-lefland18}{Lefland et al., 2018})}.
Overall, while edge effects near degraded habitat edges are often negative, meta-population \emph{(\protect\hyperlink{ref-levins79}{Levins, 1979})} and -community \emph{(\protect\hyperlink{ref-leibold04}{Leibold et al., 2004}; \protect\hyperlink{ref-warren15}{Warren et al., 2015})} theories highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity.
Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests \emph{(\protect\hyperlink{ref-oliveira-neto17}{Oliveira-Neto et al., 2017})}, forests affected by logging \emph{(\protect\hyperlink{ref-clark12}{Clark and Covey, 2012}; \protect\hyperlink{ref-edwards14}{Edwards et al., 2014})}, and timber plantations \emph{(\protect\hyperlink{ref-pryde15}{Pryde et al., 2015})}.

This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade that would suppress less shade-tolerant pioneer tree taxa and allow for more tree diversity closer to the primary forest edge.
Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

@@

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center (approx. 8.3778176, -83.2935426) near the Piro Biological Station run by Osa Conservation.
(See \emph{Taylor et al. (\protect\hyperlink{ref-taylor15}{2015})} for a broader ecosystem description of the region.)
The study site was a regrowing 20 ha timber plantation of \emph{Bombacompsis quinata} abandoned in \textasciitilde1990 after the vulnerably dry-adapted species from the Guanacaste region \emph{(\protect\hyperlink{ref-hulshof20}{Hulshof and Powers, 2020}; \protect\hyperlink{ref-perezcordero02}{Pérez Cordero and Kanninen, 2002})} grew poorly in very wet conditions.
This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \ref{fig:map}) and a wide service road on the third NE border, with primary forest beyond it.
This census was done in 2013 during the rainy season months between June and August.

\begin{figure}
\centering
\includegraphics{merge_files/figure-latex/map-1.pdf}
\caption{\label{fig:map}Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica. Map image data © 2009 Google, © 2022 Osa Conservation. Photo credit: Nicholas Medina.}
\end{figure}

@@

Light reaching the forest floor was measured at the center of each plot at chest height using a densiometer \emph{(Forestry Suppliers, Inc.)}, as an average of four readings taken facing each cardinal direction.

The slope of the forest floor was measured using a rangefinder \emph{(Bushnell, Forestry Suppliers, Inc)} to measure the distance the diagonal between two plot corners and triangulate the observation angle.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage[left]{lineno}
\linenumbers
\usepackage{setspace}\doublespace
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a reforested wet neotropical timber plantation},
  pdfauthor={Nicholas Medina * a,d; Emma Capetz c; Andrea Johnson b; Agustín Mendoza b; Max Villalobos b},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Habitat edges affect tree diversity more than biomass regeneration in a reforested wet neotropical timber plantation}
\author{Nicholas Medina * \textsuperscript{a,d} \and Emma Capetz \textsuperscript{c} \and Andrea Johnson \textsuperscript{b} \and Agustín Mendoza \textsuperscript{b} \and Max Villalobos \textsuperscript{b}}
\date{\scriptsize \textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA 02453 USA; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Puntarenas, Costa Rica; \textsuperscript{c} Lehigh University, 27 Memorial Drive West, Bethlehem, PA 18015 USA \textsuperscript{d} Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI 48109 USA}

@@

*\textbf{Correspondence}: Nicholas Medina, \href{mailto:nmedina@umich.edu}{\nolinkurl{nmedina@umich.edu}}

@@

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests.
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely.
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration.
However, most studies assess how degraded edges affect intact forests, leaving it unclear whether higher-quality edges could facilitate regeneration of nearby degraded forests.
Notably, higher quality edges near intact forests could promote processes like dispersal and wood biomass accumulation that effectively accelerate succession, leading to better performance of shade-tolerant taxa compared to pioneer taxa in the early stages of forested plantation regeneration.
This study addressed how wet tropical forested plantation regeneration was affected by distance to adjacent intact forest edge.
It was hypothesized that old timber plantations facilitate regeneration by increasing available shade, favoring the presence and biomass of later-successional taxa, ultimately changing community composition overall.
A wet neotropical timber plantation reforested after 20 years and adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched identified taxa to broad dispersal mode and wood density traits using relevant literature.

As distance from primary forest edge increased, stem and wood density tended to increase significantly, with \textasciitilde10\% variation explained, while biomass and canopy light surprisingly tended to stay the same.
Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, explained in part by wood density, and taxonomic composition varied notably.
Finally, tree taxa associated with both early and late successional stages decreased significantly, as well as genus Ficus, but biomass by dispersal mode did not tend to change.
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

@@

\hypertarget{resumen}{%
\section*{Resumen}\label{resumen}}
\addcontentsline{toc}{section}{Resumen}

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques quedan entre un km de su borde, efectos de borde probablemente afectan la regeneración de composición de comunidad.
Sin embargo, la mayoría de estudios evalúan cómo bordes ya degradados afectan a bosques intactos, dejándolo poco claro si bordes de más alta calidad podrían facilitar la regeneración de bosques cercanos previamente degradados.
Notablemente, bordes de alta calidad cerca de bosques intactos podrían promover procesos como disperso y acumulación de biomasa maderosa que efectivamente acceleran a la sucesión, permitiendo mejor rendimiento de taxa tolerantes a sombra comparado con taxa más adaptados a alta luz en las etapas tempranas de regeneración de plantaciones foresterados.
Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.

Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10\% variación explicado, mientras tanto biomasa y luz de dócel asombrosamente tendió a quedarse igual.
Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, explicado en parte por densidad de madera, y composición taxonomica varió notablemente.
Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, tanto como el género Ficus, pero biomasa por modo de disperso no tendió a cambiar.
Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.

@@

Forest and landscape restoration is a key international conservation and climate change adaptation strategy \emph{(\protect\hyperlink{ref-depinto20}{De Pinto et al., 2020})}.
While tropical forests specifically store most land biomass \emph{(\protect\hyperlink{ref-pan13}{Pan et al., 2013})}, most are now secondary \emph{(\protect\hyperlink{ref-fao20}{FAO, 2020})} and functionally degraded \emph{(\protect\hyperlink{ref-hubau20}{Hubau et al., 2020})}.
This likely amplifies amid increasing climate stressors \emph{(\protect\hyperlink{ref-anderegg22}{Anderegg et al., 2022})} and collective management issues, including insufficient policy support \emph{(\protect\hyperlink{ref-chazdon18}{Chazdon, 2018})} via skewed priorities \emph{(\protect\hyperlink{ref-pyron22}{Pyron and Mooers, 2022})}.
Overall, secondary forests can regenerate relatively quickly compared to old-growth forests, in some ways.
Neotropical secondary forests can grow quickly enough to accumulate biomass \textgreater10x faster than old-growth forests \emph{(\protect\hyperlink{ref-finegan96}{Finegan, 1996}; \protect\hyperlink{ref-poorter16}{Poorter et al., 2016})}, and substantially offset carbon emissions, with estimates ranging from \textasciitilde10\% from Amazonian deforestation \emph{(\protect\hyperlink{ref-smith20}{Smith et al., 2020})} to a decade of fossil fuel emissions from all of Latin American and the Caribbean \emph{(\protect\hyperlink{ref-chazdon16}{Chazdon et al., 2016})}.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80\% of old-growth forest levels in just 20 years, especially in wetter regions \emph{(\protect\hyperlink{ref-rozendaal19}{Rozendaal et al., 2019})}.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services \emph{(\protect\hyperlink{ref-guariguata01}{Guariguata and Ostertag, 2001}; \protect\hyperlink{ref-stanturf14}{Stanturf et al., 2014})} like biomass and carbon storage \emph{(\protect\hyperlink{ref-liu18}{Liu et al., 2018})}.
However, recovery of secondary neotropical forests diverge widely in taxonomic composition \emph{(\protect\hyperlink{ref-norden15}{Norden et al., 2015})}, potentially taking over a century to recover \emph{(\protect\hyperlink{ref-poorter21}{Poorter et al., 2021})} with added variability \emph{(\protect\hyperlink{ref-atkinson22}{Atkinson et al., 2022})}.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa \emph{(\protect\hyperlink{ref-carlo16}{Carlo and Morales, 2016})}, which can be keystone to locally-adapted food webs, including birds \emph{(\protect\hyperlink{ref-maas16}{Maas et al., 2016})}.
Overall, understanding secondary forest regeneration is key for global biodiversity conservation at the global scale.

Modern secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, especially in heterogeneously-managed landscapes \emph{(\protect\hyperlink{ref-melo13}{Melo et al., 2013}; \protect\hyperlink{ref-perfecto09}{Perfecto et al., 2009})}.
About 70\% of all forests have been estimated to be within a km of their edge \emph{(\protect\hyperlink{ref-haddad15}{Haddad et al., 2015})}, resulting in tropical forests having been estimated to lose \textasciitilde22\% aboveground carbon along heavily-managed edges \textgreater100m, in part due to shortening trees and lighter leaf mass \emph{(\protect\hyperlink{ref-ordway20}{Ordway and Asner, 2020})}.
Effects of degraded edges have historically favored tree taxa that tend to be pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant \emph{(\protect\hyperlink{ref-tabarelli08}{Tabarelli et al., 2008})}.
Accordingly, secondary forest wood density may also tend to decrease near degraded habitat edges.
Additionally, the dominance of pioneering taxa could lead to arrested succession \emph{(\protect\hyperlink{ref-tymen16}{Tymen et al., 2016})}, which further delays the recruitment of shade-tolerant taxa, as well as associated increases in aboveground biomass, in part due to their higher survival with increasing shade, and tendency toward relatively higher wood densities.
However, arrested succession could be prevented, and even typical succession accelerated, if intact forests that were adjacent to regenerating plantations functioned as contributing reservoirs of seeds from later-successional tree taxa.
In this case, having a habitat edge near a relatively rich intact forest, rather than near a relatively degraded open area, would instead be an asset to regenerating plantation forest patches, rather than a detriment.
Overall, edge effects ultimately have the potential to shape community composition, in part based on the local dispersal kernel patterns of existing taxa \emph{(\protect\hyperlink{ref-wandrag17}{Wandrag et al., 2017})}, which while complicated \emph{(\protect\hyperlink{ref-muller-landau09}{Muller-Landau and Hardesty, 2009})}, can also correlate with simpler plant traits like height and seed mass \emph{(\protect\hyperlink{ref-tamme14}{Tamme et al., 2014}; \protect\hyperlink{ref-thomson11}{Thomson et al., 2011})}.

Additionally, properties of the tropical forest carbon stores and tree community depend heavily on land-use history and previous management \emph{(\protect\hyperlink{ref-omeja12}{Omeja et al., 2012}; \protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}.
Overall, legacy effects from management like intensive timber planting could potentially delay or result in non-linear changes between successional stages \emph{(\protect\hyperlink{ref-albrich20}{Albrich et al., 2020}; \protect\hyperlink{ref-gough21}{Gough et al., 2021})}, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry \emph{(\protect\hyperlink{ref-lefland18}{Lefland et al., 2018})}.
In some cases, the leftover timber stands could provide additional shade, which would not be available in a clear-cut harvested timber plantation, and this could help more shade-tolerant taxa establish and recruit better.
Overall, edge effects near degraded habitat edges are often studied as negative, but conceptual influences from meta-population \emph{(\protect\hyperlink{ref-levins79}{Levins, 1979})} and meta-community \emph{(\protect\hyperlink{ref-leibold04}{Leibold et al., 2004}; \protect\hyperlink{ref-warren15}{Warren et al., 2015})} theories do highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity, thereby potentially making particular edge effects, specifically from adjacent intact forests, actually beneficial for regeneration plantation forest parcels.
Taken together, these ideas suggest that edge effects specifically from adjacent intact primary forests, potentially via shade-tolerant dispersal and performance, may indeed facilitate plantation forest regeneration.
Such a beneficial edge effect may ultimately help explain relatively high biodiversity, compared to adjacent stands, found in agroforests \emph{(\protect\hyperlink{ref-oliveira-neto17}{Oliveira-Neto et al., 2017})}, forests affected by logging \emph{(\protect\hyperlink{ref-clark12}{Clark and Covey, 2012}; \protect\hyperlink{ref-edwards14}{Edwards et al., 2014})}, and timber plantations \emph{(\protect\hyperlink{ref-pryde15}{Pryde et al., 2015})}.

Influenced by restoration ecology for forest management, this study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, focused on highlighting edge effects from an adjacent primary forest on overall stand properties, tree community composition, and limited functional trait recovery.
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade, which would suppress less shade-tolerant pioneer tree taxa, and in turn facilitate more shade-tolerant tree taxa, ultimately allowing for more tree diversity within the regenerating plantation forest in parts that were closer to the richer edge of the intact unmanaged forest.
Accordingly, we predicted more specifically that further from the intact primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

@@

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center (approx. 8.3778176, -83.2935426) near the Piro Biological Station run by Osa Conservation.
(See \emph{Taylor et al. (\protect\hyperlink{ref-taylor15}{2015})} for a broader ecosystem description of the region.)
The study site was a regrowing 20 ha timber plantation of \emph{Bombacompsis quinata} abandoned in \textasciitilde1990 after the vulnerably dry-adapted species from the Guanacaste region
\emph{(\protect\hyperlink{ref-hulshof20}{Hulshof and Powers, 2020})} grew poorly in very wet conditions.
This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \ref{fig:map}) and a wide service road on the third NE border, with primary forest beyond it.
This census was done in 2013 during the rainy season months between June and August.

\begin{figure}
\centering
\includegraphics{merge_files/figure-latex/map-1.pdf}
\caption{\label{fig:map}Map of (a) conservation areas and (b) study site (c) showing plot design with 50 m distance strata bins, and (d) focal plantation timber taxon Bombacopsis quinata on Osa Peninsula, Costa Rica. Map image data © 2022 Osa Conservation, © 2009 Google. Map GIS data and photo credit: Max Villalobos, Nicholas Medina.}
\end{figure}

@@

Light reaching the forest floor was measured at the center of each plot at chest height using a densiometer \emph{(Forestry Suppliers, Inc.)}, as an average of four readings taken facing each cardinal direction.
The slope of the forest floor was measured using a rangefinder \emph{(Bushnell, Forestry Suppliers, Inc)} to measure the distance the diagonal between two plot corners and triangulate the observation angle.

The diameter of all stems \textgreater10 cm wide were recorded in each census plot, totaling over 1,243 trees.
In cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
Tree height was recorded by measuring distances to both the crown and the stem at chest height \emph{(\textasciitilde2.7 m)} using a rangefinder \emph{(Bushnell, Forestry Suppliers, Inc.)} and triangulating the missing side length.
